---
title: "PM 516B Project Report"
author:
- Brandyn Ruiz
- Yichi Zhang
date: "08/10/2021"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(psych)
library(tidyverse)
library(modelsummary)
library(survival)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(survminer)
library(mfp)
library(cmprsk)
library(kableExtra)
library(forcats)
library(stringr)
library(cowplot)
library(survMisc)
library(survivalAnalysis)
```

## Executive Summary

- 
- 

## Background and Goals

This study was performed to measure if there were any disparities in patient liver transplants by a generated social vulnerability index and how long the patients survive the transplant. The patients in the dataset have liver cancer and have demographic information on age at the time of initial listing, gender, highest education level, blood type, and body mass index to name a few. The patients were measured by insurance, whether they had reliable transportation, social support, and their zip code of residence as metrics to derive the generated social vulnerability index as some patients on the live transplant waitlist are limited and do not have the necessary resources. The scope of our analysis would be to perform a survival analysis to explore the different times and lengths patients are living past their liver transplant and what demographics these patients make up stratifying by their generated social vulnerability index. As well as a competing risk analysis with those receiving a transplant and the risk of mortality while on the waitlist.

## Study Design & Data

With our data having recorded variables of initial start time for each patient when they first got on the waitlist, composite death date, and transplant date this depicts an observational cohort study design. The data enables a clinical trial where patients can be admitted to the study at any given time with these variables recorded for everyone, we are able to perform a survival analysis as each patient is followed from initial start to time of even, whether they received a liver transplant or died while on waitlist. The data is sampled from the USC Norris Cancer Center with patients having liver cancer and being admitted to the liver transplant waitlist. 

Below is the demographics of patients displayed by the created social vulnerability index. Note the social vulnerability index was divided into quantiles.

```{r, include = FALSE}
# read in dataset
data <- read_csv("sdoh-for Trevor.csv")
data <- mutate(data,
               vuln = scale(pcs_wl),  
               vuln.q2 = cut(vuln, quantile(vuln, probs=0:2/2), include.lowest=T),
               vuln.q4 = cut(vuln, quantile(vuln, probs=0:4/4), include.lowest=T),
               meld = init_meld_peld_lab_score,
               meld.q2 = cut(meld, quantile(meld, probs=0:2/2), include.lowest=T),
               meld.q4 = cut(meld, quantile(meld, probs=0:4/4), include.lowest=T),
               outcome = case_when((death == 1 & lt == 0) ~ 1, #death
                             (death == 0 & lt == 1) ~ 2, #liver transplant
                             (death == 0 & lt == 0) ~ 3),#nothing happened
               outcome.f = factor(outcome, labels=c("Censored", "Dead", "Transplant")))
data$gender <- factor(data$gender)
data$abo <- factor(data$abo)
data$diab[data$diab == 998] <- NA
data$diab <- factor(data$diab)
data$ethcat <- factor(data$ethcat)
data$insure_tcr <- factor(data$insure_tcr)
data$education[data$education == 998] <- NA
data$education <- factor(data$education)
data$waitregion <- factor(data$waitregion)
data$tumorcat4 <- factor(data$tumorcat4)
```

```{r, echo = FALSE}
demo <- data %>%
  select(gender, age, "weight" = wgt_kg_tcr, "height" = hgt_cm_tcr, "BMI" = bmi_tcr, "tumor number" = tumornum1, "tumor size" = first_tumor_diam,"tumor category" = tumorcat4, "social vulnerability index" = pcs_wl, "meld score" = meld, "waiting days" = dayswait_chron, mosexctoend, vuln.q4)
# demographics table by group
demo_tab <- datasummary_balance(all ~ vuln.q4, data = demo, dinm = FALSE, output = "data.frame")
colnames(demo_tab) <- c(" "," ", rep(c("Mean", "SD"),4))
demo_tab %>%
  kable() %>%
  add_header_above(c(" " = 2, "25%" = 2, "50%" = 2, "75%" = 2, "100%" = 2))%>%
  kable_classic(full_width = F)%>%
  kable_styling(font_size = 7)
```


## Statistical Methods

Using survival analysis models, we can obtain waitlist mortality to show the survival rates of the patients as time progresses. We also use Cox Proportional Hazards model to see which variables contribute the most to the risk of mortality while waiting on the liver transplant waitlist. We replicate the same models for those that have received a liver transplant to observe which variables contribute risk and make the patients most critical in receiving a liver transplant.

## Results

### Death

#### Cox Propotional Hazard Model

```{r Death Survival, echo=FALSE}
surv_object <- Surv(time = data$mosexctoend, event = data$death)
#head(surv_object, 20)
surv1 <- survfit(surv_object ~ 1, data = data)
surv1
#summary(surv1)
ggsurvplot(surv1, data = data, title = 'Survival Plot of Death Outcome',
           xlab = 'Time (Months)', ylab = 'Survival Probability')
surv2 <- survfit(surv_object ~ vuln.q4, data = data)
surv2
ggsurvplot(surv2, data = data, title = 'Survival Plot of Death Outcome by SVI',
           xlab = 'Time (Months)', ylab = 'Survival Probability')
surv3 <- survfit(surv_object ~ meld.q4, data = data)
surv3
ggsurvplot(surv3, data = data, title = 'Survival Plot of Death Outcome by Meld Performance',
           xlab = 'Time (Months)', ylab = 'Survival Probability')
#summary(surv3, times = seq(0, 90, 1))
# Survival table
```

```{r Death Coxph Models, echo=FALSE}
#Death
cox <- coxph(surv_object ~ gender + abo + diab + ethcat + education +
               vuln.q4 + strata(meld.q4) + first_tumor_diam + 
               tumorcat4, data = data)
#summary(cox)
cox_death <- cox_as_data_frame(cox)%>%
  select(factor.id, HR, Inv_Lower_CI, Inv_Upper_CI, p)%>%
  rename(Variable = factor.id,
         'Lower CI' = Inv_Lower_CI,
         'Upper CI' = Inv_Upper_CI,
         'p-value' = p)%>%
  kbl(caption = 'Cox Proportional Hazards for Death Outcome')%>%
  kable_styling('striped', full_width = F, html_font = 'Cambria')%>%
  row_spec(c(10:11, 20, 23, 24, 26), bold = TRUE)
cox_death
```

```{r, echo=FALSE}
cox.zph(cox)
ggcoxzph(cox.zph(cox))
```

```{r, echo=FALSE}
# Deviance Residuals ggplot
ggcoxdiagnostics(cox, type = "deviance", sline = F, ox.scale = "observation.id")+
  geom_hline(yintercept = 2, color = "orange") +
  geom_hline(yintercept = -2, color = "orange")
# Row selection of observation ouliers
table <- residuals(cox, type = "deviance")%>%
  data.frame()%>%
  tibble::rownames_to_column()%>%
  filter(abs(.) > 2)%>%
  arrange(desc(.))%>%
  head(10)
colnames(table)[2] <- 'Deviance Residuals'
table%>%
  kbl(caption = 'Observation Outliers')%>%
  kable_styling('striped', full_width = F, html_font = 'Cambria')
dev_table <- data[table$rowname, ]%>%
  select(gender, abo, ethcat, education, tumorcat4, vuln.q4, meld.q4,
         outcome, mosexctoend)%>%
  kbl(caption = 'Outliers')%>%
  kable_styling('striped', full_width = FALSE, html_font = 'Cambria')
dev_table
# ggcoxdiagnostics(cox, type = "dfbeta", sline = F)+
#   geom_hline(data = bind_rows(tibble(val = abs(coefficients(cox)*.1),
#                                      covariate = names(coefficients(cox))),
#                               tibble(val = -abs(coefficients(cox)*.1),
#                                      covariate = names(coefficients(cox)))),
#              aes(yintercept = val), color = "orange")
#^Displays every level of every covariate and is overwhelming to explore, should we week it?
```

#### Competing Risk Analysis

```{r, echo=FALSE, cache = TRUE}
death_hr_fit <- crr(ftime = data$mosexctoend,
              fstatus = data$outcome,
              cov1 = model.matrix(~ fct_explicit_na(factor(gender)) + 
                                    fct_explicit_na(factor(abo)) +
                                    fct_explicit_na(factor(diab)) +
                                    fct_explicit_na(factor(ethcat)) +
                                  fct_explicit_na(factor(education)) +
                                    vuln.q4 +
                                    first_tumor_diam +
                                    fct_explicit_na(factor(tumorcat4)),
                                  data = data)[,-1],
  cengroup = data$meld.q4,
              failcode = 2, cencode = 3)
mvcrrres <- function(mod) {
  names <- names(mod$coef)
  hr.ci <- paste0(
    round(summary(mod)$conf.int[, 1], 2), " (",
    round(summary(mod)$conf.int[, 3], 2), ", ",
    round(summary(mod)$conf.int[, 4], 2), ")"
  )
  p <- round(summary(mod)$coef[, 5], 3)
  res <- data.frame(names, hr.ci, p, stringsAsFactors = FALSE)
  res$p[res$p < 0.001] <- "<.001"
  colnames(res) <- c("Variable", "HR (95% CI)", "p-value")
  return(res)
}
death_crr <- mvcrrres(death_hr_fit)
rownames(death_crr) <- NULL
death_crr <- death_crr%>%
  filter(`p-value` < 0.05)
death_crr$Variable <- gsub("fct_explicit_na", "", death_crr$Variable)
death_crr$Variable <- gsub("[()]", "", death_crr$Variable)
death_crr$Variable <- gsub("factor", "", death_crr$Variable)
death_crr <- death_crr[!grepl("Missing", death_crr$Variable),]%>%
  kbl(caption = 'Death Competing Risks Regression')%>%
  kable_styling('striped', full_width = F, html_font = 'Cambria')
death_crr
```


### Transplant

#### Cox Propotional Hazard Model

We start with the following set of covariates to build the cox proportional hazards model: gender, ethnicity, diabetes, BMI, tumor diameter, tumor number, tumor category, meld score and the created vulnerability index. However, after examining the proportional hazards assumption, we decided to remove tumor diameter and tumor number since they are significantly correlated with tumor category. We then removed ethnicity and BMI because they are not significantly related to overall survival and they violate the proportional hazards assumption. MELD score is stratified in this model. We use **tt** function to address non-proportional hazards of tumor category and vulnerability index and this model has a concordance of 0.534. The table shows the coefficients for covariates in the cox hazard model with bolded line indicating significant effect. For example, being male increases the hazard of transplant by a factor of 1.068, or 6.8%.

```{r, echo = FALSE}
surv_trans <- Surv(time = data$mosexctoend, event = data$lt)
cox3.m <- coxph(surv_trans ~ diab + gender + tumorcat4
                + vuln + strata(meld.q4), data = data)
test.ph <- cox.zph(cox3.m)
test.ph
res.cox <- coxph(Surv(time = data$mosexctoend, event = data$lt) ~ diab + gender + tt(tumorcat4) + tt(vuln.q4) + strata(meld.q4),
                  data = data,
                  tt = list(
                    function(tumorcat4, mosextoend, ...){
                      mtrx <- model.matrix(~tumorcat4)[,-1]
                      mtrx * mosextoend
                    },
                    function(vuln.q4, mosextoend, ...){
                      mtrx <- model.matrix(~vuln.q4)[,-1]
                      mtrx * mosextoend
                    }
                  ))
# summary(res.cox)
cox_trans <- cox_as_data_frame(res.cox)%>%
  select(factor.id, HR, Inv_Lower_CI, Inv_Upper_CI, p)%>%
  rename(Variable = factor.id,
         'Lower CI' = Inv_Lower_CI,
         'Upper CI' = Inv_Upper_CI,
         'p-value' = p)%>%
  kbl(caption = 'Cox Proportional Hazards for Transplant Outcome')%>%
  kable_styling('striped', full_width = F, html_font = 'Cambria')%>%
  row_spec(c(2,3,5, 10, 11), bold = TRUE)
cox_trans
```

```{r, include=FALSE, cache=TRUE}
# Deviance Residuals ggplot
# ggcoxdiagnostics(res.cox, type = "deviance", sline = F, ox.scale = "observation.id")+
#   geom_hline(yintercept = 2, color = "orange") +
#   geom_hline(yintercept = -2, color = "orange")
# # # Row selection of observation ouliers
# table_tran <- residuals(res.cox, type = "deviance")%>%
#   data.frame()%>%
#   tibble::rownames_to_column()%>%
#   filter(abs(.) > 2)%>%
#   arrange(desc(.))%>%
#   head(10)
# colnames(table_tran)[2] <- 'Deviance Residuals'
# table_tran%>%
#   kbl(caption = 'Observation Outliers')%>%
#   kable_styling('striped', full_width = F, html_font = 'Cambria')
# dev_table_tran <- data[table$rowname, ]%>%
#   select(gender, abo, ethcat, education, tumorcat4, vuln.q4, meld.q4,
#          outcome, mosexctoend)%>%
#   kbl(caption = 'Outliers')%>%
#   kable_styling('striped', full_width = FALSE, html_font = 'Cambria')
# dev_table_tran
# ggcoxdiagnostics(cox, type = "dfbeta", sline = F)+
#   geom_hline(data = bind_rows(tibble(val = abs(coefficients(cox)*.1),
#                                      covariate = names(coefficients(cox))),
#                               tibble(val = -abs(coefficients(cox)*.1),
#                                      covariate = names(coefficients(cox)))),
#              aes(yintercept = val), color = "orange")
#^Displays every level of every covariate and is overwhelming to explore, should we week it?
```

Below is the survival plot by quantiles of the social vulnerability index. The log-rank test suggests three is a statistically significant difference in survival curves between quantiles of social vulnerability index ($p < 0.0001$).

```{r, echo = FALSE, fig.align='center'}
# surv_trans
surv4.m <- survfit(surv_trans ~ vuln.q4, data=data)
# summary(surv4.m)
ggsurvplot(surv4.m, data = data, title = 'Survival Plot of Transplant Outcome by SVI',
           xlab = 'Time (Months)', ylab = 'Survival Probability', pval = T)
survdiff(surv_trans ~ vuln.q4, data = data)
surv5.m <- survfit(surv_trans ~ meld.q4, data = data)
ggsurvplot(surv5.m, data = data, title = 'Survival Plot of Transplant Outcome by Meld Performance', xlab = 'Time (Months)', ylab = 'Survival Probability', pval = T)
```


#### Competing Risk Analysis

Below is the cumulative incidence plot for each outcome, by dichotomized meld score. It suggests patients with higher vs. lower MELD scores are more likely to get a transplant early, but then lower vs. higher MELD scores are more likely to get a transplant after about 18 months. Again, the proportional odds assumption doesn't hold for MELD score. 

```{r, echo = FALSE}
plot(cuminc(data$mosexctoend, 
            data$outcome.f,
            data$meld.q2,
            cencode=0),
     color=c(1, 1, 2, 2, 3, 3))
```

Thus, we estimate different censoring distributions based on MELD score group in the competing risks regression model. The results suggests, patients with higher vulnerability index is associated with lower hazard of getting a transplant.

```{r, echo = FALSE, cache=TRUE, warning = FALSE}
scho.plot <- function(m) {
  require(magrittr)
  
tibble(m$uft) %>%
  bind_cols(as_tibble(m$res)) %>%
  set_colnames(
    c("ft", names(m$coef))
  ) %>%
  pivot_longer(
    cols = names(m$coef),
    names_to = "var",
    values_to = "resid"
  ) %>%
  ggplot(aes(x=ft, y=resid)) +
  facet_wrap(~var) +
  geom_point(alpha=.2) +
  geom_smooth(color="firebrick")
}
# diab + gender + tumorcat4 + vuln
result5b1 <- crr(data$mosexctoend,
                data$outcome,
                cbind(
                  model.matrix(~ fct_explicit_na(factor(data$vuln.q4)))[,-1],
                  model.matrix(~ fct_explicit_na(factor(data$gender)))[,-1],
                  model.matrix(~ fct_explicit_na(factor(data$diab)))[,-1],
                  model.matrix(~ fct_explicit_na(factor(data$tumorcat4)))[,-1]
                ),
                cengroup = data$meld.q4,
                failcode=2, cencode=0)
# summary(result5b1)
scho.plot(result5b1)
trans_crr <- mvcrrres(result5b1)
rownames(trans_crr) <- NULL
trans_crr <- trans_crr%>%
  filter(`p-value` < 0.05)
trans_crr$Variable <- gsub("fct_explicit_na", "", trans_crr$Variable)
trans_crr$Variable <- gsub("[()]", "",trans_crr$Variable)
trans_crr$Variable <- gsub("factor", "", trans_crr$Variable)
trans_crr[!grepl("Missing", trans_crr$Variable),]%>%
  kbl(caption = 'Transplant Competing Risks Regression')%>%
  kable_styling('striped', full_width = F, html_font = 'Cambria')
trans_crr 
```

